\section{Linear Functions Review Questions}

\subsection*{Slope}

\bigskip
\ex{$g(x)$ is a linear function with $g(-1) = 5$ and $g(4) = -6$. Find its slope.}
\sol{$-2.2$}

\bigskip
\ex{$f(x)$ is a linear function with $f(-5) = -14$ and $f(1) = -2$. Find its slope.}
\sol{$2$}

\bigskip
\ex{$g(t)$ is a linear function with $g(2) = -19$ and $g(8) = -49$. Find its slope.}
\sol{$-5$}

\bigskip
\ex{Suppose $k(m)$ is linear. Use the table below to find its slope.
	$$
	\begin{array}{c|c}
		m & k(m) \\
		\hline
		-7 & 3,000 \\
		10 & -400 \\
		250 & -48,400 \\
	\end{array}
	$$
}
\sol{$-200$}

\bigskip
\ex{Given the table below, find the slope of the linear function.
	$$
	\begin{array}{c|c}
		x & f(x) \\
		\hline
		-2 & 6 \\
		2 & -4 \\
		7 & -16.5 \\
		10 & -24 \\
	\end{array}
	$$
}
\sol{$-2.5$}

\bigskip
\ex{Find the slope of the linear function whose graph passes through the points $(4,1)$ and $(10, -8)$.}
\sol{$-1.5$}

\bigskip
\ex{Find the slope of the linear function whose graph passes through the points $(0, -12)$ and $(14, -8.5)$.}
\sol{$0.25$}

\subsection*{Intercepts}

\bigskip
\ex{Determine the vertical and horizontal intercepts for each of the linear functions.
	\begin{enumerate}[label=(\alph*)]
		\item $g(t) = -5t + 20$
		\item $f(q) = 16 - \frac{3}{4}q$
		\item $h(k) = -1.2(k - 3) + 4$
		\item $j = \frac{2x}{5}$
		\item $h(t) = -2t + 4$
		\item $g(n) = 2n - 7$
		\item $f(b) = \frac{1}{2}b + \frac{7}{3}$
	\end{enumerate}
}
\sol{a.  Vertical: $(0,20)$, Horizontal: $(4,0)$\\\b.  Vertical: $(0,16)$, Horizontal: $(\frac{64}{3},0)=(21.33,0)$\\c.  Vertical: $(0,7.6)$, Horizontal: $(\frac{19}{3},0)=(6.33,0)$\\d.  Vertical: $(0,0)$, Horizontal: $(0,0)$\\e.  Vertical: $(0,-7)$, Horizontal: $(3.5,0)$\\f.  Vertical: $(0,\frac{7}{3})=(0, 2.33)$, Horizontal: $(-\frac{14}{3},0)=(-4.67,0)$
}

\bigskip
\ex{If the function is linear, state the slope and vertical intercept.
	\begin{enumerate}
		\item $a(b) = 34 - 50b$
		\item $c = 10 - 2(x + 4)$
		\item $d(f) = 15(f - 2) + 4(2f + 6)$
		\item $h(x) = 2(x - 2) + x(x + 4)$
	\end{enumerate}
}
\sol{a.  Slope: $-50$, Vertical: $(0,34)$\\b.  Slope: $-2$, Vertical: $(0,2)$\\c.  Slope: $23$, Vertical: $(0,-6)$\\d.  Not linear}

\bigskip
\ex{The number of bags of fertilizer needed to fertilize a lawn of size $A$ acres is given by the function $f(A) = 11A$. Write a sentence interpreting the vertical intercept and slope.
}
\sol{Slope:  The number of bags of fertilizer needed increases by $11$ bags per acre.  Vertical: No bags of fertilizer are needed for zero acres.}

\bigskip
\ex{The monthly water bill for a resident of the town of Williamson is given by the function $w(g) = 25 + 0.005g$. Write a sentence interpreting the vertical intercept and slope.
}
\sol{Slope:  The residents of the town of Williamson are charged an additional $\$0.005$ per additional gallon of water used each month.  Vertical: The residents of the town of Williamson are charged a monthly fee of $\$25$ for water.}

\bigskip
\ex{\label{LinearFunctionsReviewExercise5}
	The cost for a cab ride is a function of the distance traveled. It costs $25.50$ to travel $15$ miles and $34.50$ to travel $21$ miles. Find the slope and vertical intercept and interpret both given the context of this function.
}
\sol{Slope:  The cost for a cab ride increase by$\$1.50$ for each additional mile.  Vertical:  It costs $\$3.00$ to have the cab pick you up.}

\bigskip
\ex{For the question above (\ref{LinearFunctionsReviewExercise5}), find the horizontal intercept. Does that have a realistic meaning here?}
\sol{$(-2,0)$; no}

\subsection*{Formula}

\bigskip
\ex{Find the equation of a linear function whose slope is $-\frac{2}{9}$ with a vertical intercept of $(0,-5)$.}
\sol{$y=-\frac{2}{9}x-5$}

\bigskip
\ex{Find the formula of a linear function whose slope is $5$ and its graph passes through the point $f(6) = 15$.}
\sol{$y=5x-15$}

\bigskip
\ex{Suppose $h(x)$ is linear, and its graph passes through the points $(-5, 0.2)$ and $(3, 10)$. Find a formula for this function.}
\sol{$y=1.225x+6.325$}

\bigskip
\ex{Find a formula for the linear function whose slope is $6$ and passes through the point $(0, -10)$.}
\sol{$y=6x-10$}

\bigskip
\ex{Find a formula for the linear function whose graph passes through the points $(-3,24)$ and $(3,0)$.}
\sol{$y=-4x+12$}

\bigskip
\ex{Given that $w(t)$ is linear and, $w(-5) = -85$ and $w(10) = 95$, find a formula for this function.}
\sol{$w(t)=12t-25$}

\bigskip
\ex{The price of a first class stamp in 1971 was $0.07$. By 2018 it rose to $0.50$. Assuming the price of a stamp is a linear function of years since 1971, write an equation that models this relationship.}
\sol{$y=0.009t+0.07$}

\bigskip
\ex{The cost to refinish a hardwood floor is a linear function of the size of the room, in square feet. It costs $\$700$ to refinish a room $100$ sq. feet. And $\$950$ for a room $150$ sq. feet. Find a formula for this linear function.}
\sol{$r(s)=5s+200$}

\subsection*{Graphs}

Solve the following equations graphically and algebraically.

\bigskip
\ex{$2x - 5 = 7x + 6$}
\sol{$4=-5.5$}

\bigskip
\ex{$-3(y - 5) = 6(7 - 2y)$}
\sol{$y=3$}

\bigskip
\ex{$1000 + 50z = 25(5z + 120)$}
\sol{$z=-\frac{80}{3} \approx -26.67$}

\bigskip
\ex{$5x - 7 = 6x + 1$}
\sol{$x=-8$}

\bigskip
\ex{$2(4 - x) + 1 = 7 + 3(-x + 5)$}
\sol{$x=13$}

\subsection*{System of Linear Equations}

\bigskip
Solve the system of equations using an algebraic method.

\bigskip
\ex{$\begin{cases} 6x - 2y = 12 \\ 2x + 2y = 36 \end{cases}$}
\sol{$x=6$ and $y=12$}

\bigskip
\ex{$\begin{cases} 3a = 7 + b \\ 2a + b = 8 \end{cases}$}
\sol{$a=3$ and $b=2$}

\bigskip
\ex{$\begin{cases} 5c + 8d = -16 \\ 17c - 2d = 4 \end{cases}$}
\sol{$c=0$ and $d=-2$}

\bigskip
\ex{$\begin{cases} 5f - 7g = 89 \\ 7g = -8 + 5f  \end{cases}$}
\sol{No Solution}

\bigskip
\ex{$\begin{cases} 20x - 6y = 8 \\ 8x - 2y = 4 \end{cases}$}
\sol{$x=1$ and $y=2$}

\bigskip
\ex{$\begin{cases} -3x + 15y = 33 \\ 3x - 6y = -6 \end{cases}$}
\sol{$x=4$ and $y=3$}

\bigskip
\ex{$\begin{cases} 10x - 4y = 20 \\ 2x - 2y = -2 \end{cases}$}
\sol{$x=4$ and $y=5$}

\bigskip
\noindent Solve the system of equations using graphically.
\bigskip
\ex{$\begin{cases} x + y = 7 \\	x + 2y = 11 \end{cases}$}
\sol{$x=3$ and $y=4$}

\bigskip
\ex{$\begin{cases} 2z + 3a = 5 \\ z + a = 4 \end{cases}$}
\sol{$z=7$ and $a=-3$}

\clearpage
